Multiple Sclerosis Diagnosis from Natural Language Processing of Patient Text Data

### 2022-09-14 paper.tex
Added footnotes

% This is not strictly necessary, and may be commented out,
% but it will improve the layout of the manuscript,
% and will typically save some space.
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{array}
\usepackage{multirow}
\newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}

@@

\begin{document}
\maketitle
\begin{abstract}
\end{abstract}

@@

\textbf{Diagnosed group.} To get textual comments outlining the experiences of MS patients, we will utilize free text data from the MS Register database \citep{Ford:12}. We chose to use this database because it provides a large, rich resource of experiences of patients diagnosed with MS, in text format. However, one problem of using this dataset directly for our use-case is that the experiences of the patients are recorded post MS diagnosis, which means patients often refer to their condition (that is, multiple sclerosis), in their comments. However, since the classifier will be used pre-diagnosis, we cannot make this classifier rely on the term “multiple sclerosis”, or any other term that only MS patients will know about. Therefore, we remove all clauses containing MS-related terms (footnote with list link) from patient comments. If every clause in a patient comment includes some MS-related term, then we omit the comment from our dataset. With this approach, we were able to get 6121 comments from 3037 diagnosed users in our dataset. \\

@@

\indent The first control group constitutes random users from Twitter and Reddit who have not mentioned the term “MS” or any related terms in their timeline. The dataset was constructed such that there is an equal probability for a potential control user to be selected from Twitter or from Reddit. Moreover, we ignored image-only posts, video-only posts, Twitter retweets and Reddit cross-posts. For Reddit, we added the additional restriction that the potential control group user has never posted on any MS-related subreddit (footnote). After identifying the users, we select random posts in their timeline to be included in our dataset. The number of posts we select per user is controlled such that the final user-comment distribution is kept the same as the diagnosed users in the dataset, as illustrated in Figure \ref{fig:usercomment}. Finally, as a measure to prevent posts created by bots to be included in our dataset, no more than 5 duplicate posts were allowed by the same user. \\

@@

\indent Therefore, the second control group are comments of users who have given self-reported diagnoses of diabetes on social media (both type 1 and type 2), and who, like the first control group, have not mentioned “MS” or any related terms in their timeline. Diabetes was chosen as the illness for the second control group because firstly, it is an incurable, chronic condition like MS, and secondly, it is relatively prevalent, so we can find enough users on social media to exactly match the number of MS-diagnosed users from the MS Register dataset. The users were found by searching on Twitter and Reddit for high-precision diagnostic patterns (footnote). After identifying these users, we only include posts where the user mentions their condition (that is, diabetes), because we would expect only those experiences to be presented to a primary healthcare provider. Then, like what we did for the diagnosed users, we remove all clauses with the word “diabetes” from the text, since the classifier will be used pre-diagnosis, and we cannot make the classifier rely on the term “diabetes” for classifying patients who do not have MS. \\

@@

\indent However, before conducting any experiments, we pre-processed the dataset as follows. We removed any punctuation and URLs from the free text and made all the characters lowercase. To handle hashtags, tags, and subreddit names in the text collected from social media, we removed the characters ‘\#’, ‘@’ and ‘r/’ from the start of every word in the free text, if the word started with any of those characters. Finally, we tokenized, lemmatized and removed stop words (footnote) from the text in the database. \\

>>>>

% This is not strictly necessary, and may be commented out,
% but it will improve the layout of the manuscript,
% and will typically save some space.
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{array}
\usepackage{multirow}
\usepackage[compact]{titlesec}
\newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}
\interfootnotelinepenalty=10000

@@

\begin{document}
\maketitle
\begin{abstract}
Multiple Sclerosis (MS) is an important and growing concern for public health. Therefore, there is a need for inexpensive yet effective methods for diagnosis of MS. One such diagnosis method involves NLP-based predictive classifiers, which take as input patient experiences in text format. However, in order to employ NLP for MS diagnosis, we require a large-scale, high-quality, labelled dataset containing experiences of MS patients and individuals who do not have MS. In this paper, we introduce the largest ever dataset for MS diagnosis from patient free text data, created by combining social media text data with free text data from the MS Register database, a database of clinically diagnosed MS patients in the UK. We examine differences in language between the MS-diagnosed group and the control group, and we explore text classification methods to identify individuals with MS.
\end{abstract}

@@

\textbf{Diagnosed group.} To get textual comments outlining the experiences of MS patients, we will utilize free text data from the MS Register database \citep{Ford:12}. We chose to use this database because it provides a large, rich resource of experiences of patients diagnosed with MS, in text format. However, one problem of using this dataset directly for our use-case is that the experiences of the patients are recorded post MS diagnosis, which means patients often refer to their condition (that is, multiple sclerosis), in their comments. However, since the classifier will be used pre-diagnosis, we cannot make this classifier rely on the term “multiple sclerosis”, or any other term that only MS patients will know about. Therefore, we remove all clauses containing MS-related terms\footnote{\url{https://www.doc.ic.ac.uk/~dj321/msnlp/msterms.txt}} from patient comments. If every clause in a patient comment includes some MS-related term, then we omit the comment from our dataset. With this approach, we were able to get 6121 comments from 3037 diagnosed users in our dataset. \\

@@

\indent The first control group constitutes random users from Twitter and Reddit who have not mentioned the term “MS” or any related terms in their timeline. The dataset was constructed such that there is an equal probability for a potential control user to be selected from Twitter or from Reddit. Moreover, we ignored image-only posts, video-only posts, Twitter retweets and Reddit cross-posts. For Reddit, we added the additional restriction that the potential control group user has never posted on any MS-related subreddit \footnote{\url{https://www.doc.ic.ac.uk/~dj321/msnlp/mssubreddits.txt}}. After identifying the users, we select random posts in their timeline to be included in our dataset. The number of posts we select per user is controlled such that the final user-comment distribution is kept the same as the diagnosed users in the dataset, as illustrated in Figure \ref{fig:usercomment}. Finally, as a measure to prevent posts created by bots to be included in our dataset, no more than 5 duplicate posts were allowed by the same user. \\

@@

\indent Therefore, the second control group are comments of users who have given self-reported diagnoses of diabetes on social media (both type 1 and type 2), and who, like the first control group, have not mentioned “MS” or any related terms in their timeline. Diabetes was chosen as the illness for the second control group because firstly, it is an incurable, chronic condition like MS, and secondly, it is relatively prevalent, so we can find enough users on social media to exactly match the number of MS-diagnosed users from the MS Register dataset. The users were found by searching on Twitter and Reddit for high-precision diagnostic patterns \footnote{List of diagnosis patterns (\url{https://www.doc.ic.ac.uk/~dj321/msnlp/diagnosispatterns.txt}) was inspired by the work of \citet{Cohan:18} on NLP-based diagnosis of mental health disorders}. After identifying these users, we only include posts where the user mentions their condition (that is, diabetes), because we would expect only those experiences to be presented to a primary healthcare provider. Then, like what we did for the diagnosed users, we remove all clauses with the word “diabetes” from the text, since the classifier will be used pre-diagnosis, and we cannot make the classifier rely on the term “diabetes” for classifying patients who do not have MS. \\

@@

\indent However, before conducting any experiments, we pre-processed the dataset as follows. We removed any punctuation and URLs from the free text and made all the characters lowercase. To handle hashtags, tags, and subreddit names in the text collected from social media, we removed the characters ‘\#’, ‘@’ and ‘r/’ from the start of every word in the free text, if the word started with any of those characters. Finally, we tokenized, lemmatized and removed stop words\footnote{List of stop words (\url{https://www.doc.ic.ac.uk/~dj321/msnlp/stopwords.txt}) was inspired by NLTK's list of English stop words} from the text in the database. \\
